---
title: "Cluster Analysis on Wine Data"
output: word_document
date: "2023-12-05"
---



## Description of Data 
### The data which I chose is the Wine Data set. The dataset consists of various measurements related to different
### chemical properties which are found in a variety types of wine.The data set comprises of 178 observations, each observation in the data set corresponds
### to an each wine sample. The data set has 13 features which are used to represent the chemical components wine. There are 13 numeric variables which like
### alcohol, malicacid, totalphenols, and more.There is a categorical variable "class" which indicates the wine class.

##  The data is taken from the UCI Machine Learning Repository. 
### URL - https://archive.ics.uci.edu/dataset/109/wine


# Task 2
```{r}
library(ggplot2)
library(cluster)
#install.packages("useful")
library(useful)
#install.packages("NbClust")
library(NbClust)
library(rgl)
#install.packages("factoextra")
library(factoextra)
library(corrplot)
#install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)

```



```{r}
wine_data <- read.table("wine.data", header = FALSE, sep = ",")

# Assign column names
col_names <- c("Class", "Alcohol", "MalicAcid", "Ash", "AlcalinityOfAsh", "Magnesium", 
               "TotalPhenols", "Flavanoids", "NonflavanoidPhenols", "Proanthocyanins",
               "ColorIntensity", "Hue", "OD280_OD315", "Proline")

colnames(wine_data) <- col_names

wineScaled<- scale(wine_data)

wineScaled<- as.data.frame(wineScaled)
```

# Correlation Plot
```{r}

cormat<-cor(wineScaled)
chart.Correlation(cormat, histogram= TRUE , pch = 15)

x<-as.factor(wineScaled$Class)
levels(x)

```

# K - means
```{r}
kmeans_model2 <- kmeans(wineScaled, centers = 6, nstart = 10)

fviz_cluster(kmeans_model2, data = wineScaled)
```

## Determining the optimal k using the elbow method

```{r}
wssplot <- function(data, nc=15){                    
  
  par(mfrow=c(1,2))
  
  wss <- NULL  
  pctExp <-NULL
  
  for (k in 1:nc)
  {
    kclus <- kmeans(data, centers=k)
    wss[k] <- kclus$tot.withinss      #store the total within SSE for given k
    pctExp[k] <- 1-wss[k]/kclus$totss
  }
  
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")
  
  plot(1:nc, pctExp, type="b", xlab="Number of Clusters",
       ylab="Pct Explained")
  
  par(mfrow=c(1,1))
}

# Plot the elbow plot
wssplot(wineScaled,nc=10)

fviz_nbclust(wineScaled, kmeans, method= "silhouette")


# Determining the optimal {k}
#based on the Elbow Plot applied above, I can observe from the plot that there is a clear elbow at k=3, which indicates that there is an optimal number of clusters
optimal_k <- 3

# Perform k-means clustering
kmeans_model <- kmeans(wineScaled, centers = optimal_k, nstart = 10)
wine_data$cluster<- as.factor(kmeans_model$cluster)
table(wine_data$Class, wine_data$cluster)
# Add cluster assignments to the original data
wine_data_with_clusters <- cbind(wine_data, Cluster = as.factor(kmeans_model$cluster))

# Print the cluster assignments
print(wine_data_with_clusters$Cluster)

fviz_cluster(kmeans_model, data= wineScaled)


```

# K- Medoid
```{r}
pam_model <- pam(wineScaled, k = 3, metric ="euclidean", stand = FALSE)


# Add cluster assignments to the original data
wineData_with_clusters <- cbind(wine_data, Cluster = as.factor(pam_model$clustering))

# Print the cluster assignments
print(wineData_with_clusters$Cluster)

fviz_cluster(pam_model, data= wineScaled, geom = "point")


```

# Hierarchial Clustering


```{r}
dist<-dist(wineScaled,method = "euclidean")
hclust_model <- hclust(dist, method = "complete")

# Cut the dendrogram to get clusters
num_clusters <- 3


clusters <- cutree(hclust_model, k = num_clusters)

plot(hclust_model, cex= 0.6, hang = -1)

rect.hclust(hclust_model, k=2,border= "blue" )

wineScaled$highclust = as.factor(cutree(hclust_model,k=3))

head(wineScaled)

table(wineScaled$Class, wineScaled$highclust)

# Add cluster assignments to the original data
winedata_with_clusters <- cbind(wine_data, Cluster = as.factor(clusters))

# Print the cluster assignments
print(winedata_with_clusters$Cluster)

plot(hclust_model, cex= 0.5, hang = -1)


```



```{r}

```



```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```



```{r}

```
